\begin{abstract}
\system\ studies a narrower and more operational version of hallucination control:
evidence-grounded question answering with an explicit answer-or-\abstain\
decision.
The current implementation now reports three stages on SQuAD~2.0 with one
question and one evidence passage as input.
Stage 1 compares an answer-only baseline with an abstain-aware variant; the
abstain-aware model reduces unsupported answers on the development split from
100.00\% to 27.97\% and improves decision-aware overall $F_1$ from 42.35 to
69.40, while answerable $F_1$ decreases from 84.82 to 66.76.
Stage 2 adds a support verifier on top of the Stage 1 abstain-aware system.
The verifier reaches 79.79 support $F_1$ on the development split at a selected
support threshold of 0.25, but the resulting gated QA system changes end-to-end
metrics only marginally: overall $F_1$ moves from 69.40 to 69.42, unsupported
answer rate from 27.97\% to 27.57\%, and answerable $F_1$ from 66.76 to 66.41.
Stage 3 then calibrates the Stage 1 QA decision score and the Stage 2 support
score with scalar temperature scaling.
On the development split, QA ECE falls from 0.2456 to 0.0753 and support ECE
falls from 0.3326 to 0.1708; both fitted temperatures saturate at 5.0, which
reveals strong overconfidence in the raw scores.
These findings support the claim that explicit abstention supervision, support
verification, and post-hoc calibration are all useful intermediate controls,
but they also show that the support signal remains the weaker part of the
pipeline even after calibration.
Taken together, the first three stages are best read as evidence for the
limits of post-hoc modular tuning in this setting:
each module improves part of the behavior, but the end-to-end
answer-support-abstain trade-off remains under-specified.
That result motivates a revised next step for the paper:
compare the current modular pipeline against a jointly trained
support-constrained multi-head learner that shares the same backbone and
optimizes answer extraction, support, and abstention under explicit
unsupported-answer and over-abstention costs.
\end{abstract}
