This draft paper frames \system\ as a staged research program for
evidence-grounded question answering with controlled abstention.
The current implemented contribution now covers Stage 1, Stage 2, and Stage 3
under a controlled SQuAD~2.0 setting:
first an answer-only baseline versus an abstain-aware alternative, then a
learned support verifier on top of the abstain-aware system, and finally a
post-hoc calibration step over the QA and support scores.

The current evidence is encouraging but mixed in the right way for a staged
project.
Stage 1 substantially improves unsupported-answer behavior and decision-aware
overall performance, but it still gives up too much answerable-question
quality.
Stage 2 then shows that support verification is learnable, yet the present
gating policy is too permissive to create a large end-to-end gain.
Stage 3 confirms that this permissiveness was accompanied by serious
overconfidence in both score streams and shows that simple scalar temperature
scaling can reduce that miscalibration substantially, although support remains
the weaker signal even after calibration.
That means the project has cleared three useful bars:
abstention supervision works, support prediction is viable, and post-hoc
calibration improves the quality of the control signals.
At the same time, the results sharply motivate the next stages, especially
tighter unsupported-answer control, because the remaining question is no longer
whether confidence can be calibrated at all, but whether calibrated QA and
support signals can finally be combined into a materially better downstream
decision policy.
That motivates a more explicit next-step formulation for the paper:
use the existing modular pipeline as a strong baseline, then compare it against
a jointly trained support-constrained multi-head learner that optimizes answer
extraction, support, and abstention under explicit unsupported-answer and
over-abstention costs while keeping the same encoder family and core dataset
setup.
If that comparison succeeds under matched conditions, the contribution becomes
more than a staged tuning project;
it becomes evidence that grounded QA is better treated as a constrained
decision problem than as answer prediction followed by post-hoc fixes.
