The current project deliberately favors scope control over breadth.
That is a strength for the first paper draft, but it also creates limitations
that should be stated clearly.

\paragraph{Current limitations.}
\begin{itemize}[leftmargin=1.25em]
  \item The active implementation currently covers Stage 1, Stage 2, and
  Stage 3 only.
  \item The task uses a single provided evidence passage rather than retrieved
  evidence.
  \item The action space is intentionally narrow and does not yet cover richer
  support labels or multi-hop settings.
  \item The current results will reflect one controlled dataset setup rather
  than a broad claim about hallucination in general.
\end{itemize}

\paragraph{Intended deployment scope.}
The current approach is best interpreted as a method for high-trust,
document-bounded QA, not as a general-purpose opinion or advice generator.
Its most plausible deployment settings are ones in which the evidence source is
explicit, the answer should remain short and verifiable, and abstaining is
preferable to guessing.
If the provided evidence is incomplete, outdated, or systematically biased, the
method may still produce a well-grounded but incomplete answer, so groundedness
should not be confused with neutrality or full real-world balance.

\paragraph{How to interpret the current result.}
The current evidence supports two related but different claims.
First, the Stage 1 run shows that explicit abstention supervision and
abstention-aware evaluation can strongly reduce unsupported answers in a simple
extractive QA system.
Second, the Stage 2 run shows that support verification is learnable on top of
that abstain-aware base.
Third, the Stage 3 run shows that both the QA and support scores were indeed
overconfident and that simple post-hoc temperature scaling can reduce that
miscalibration substantially.
However, the calibrated support signal remains noticeably weaker than the QA
signal and has not yet been turned into a better end-to-end controller.
The result should therefore be read as a trade-off improvement plus two
successful intermediate controls, not yet as a generally better grounded QA
model.
The strongest interpretation is narrower:
Stages 1--3 support the usefulness of modular post-hoc controls, but they also
suggest that modular tuning alone may have limited headroom in this setup.

\paragraph{Why this stage is still useful.}
Even under its narrow scope, the first three stages now test three defensible
claims with clean supervision and interpretable failure modes.
Stage 1 shows that abstention can be taught.
Stage 2 shows that support can be predicted, but also reveals the remaining
downstream bottleneck clearly.
Stage 3 then confirms that this bottleneck was not only architectural but also
statistical:
the raw confidence scores were too sharp to trust directly.
After calibration, the QA signal becomes reasonably well behaved, while support
still remains the noisier control variable.

\paragraph{Legacy plan and revision.}
The original roadmap after Stage 1 was to move directly from Stage 2 support
verification into Stage 4 unsupported-confidence control, while treating
Stage 3 calibration as supporting work and deferring retrieval to a later
realism check.
That legacy plan assumed that once support became learnable, the raw verifier
scores would already be good enough to drive downstream control.
Stage 2 changed that assumption.
On the dev split, support recall reaches 99.43\%, but support precision is only
66.62\%, the verifier predicts support for 98.88\% of answered cases, and the
gated system improves overall \(F_1\) by only 0.03 points.
Stage 3 then validates the revised plan directly.
Both fitted temperatures saturate at 5.0, dev QA ECE drops from 0.2456 to
0.0753, and dev support ECE drops from 0.3326 to 0.1708, but support still
remains less calibrated than QA after the fix.
So the long-term goal stays the same, and the immediate plan is now clearer
than before:
Stage 4 should operate on calibrated signals rather than raw ones, because the
current support scores were too permissive before calibration and are only now
approaching the level needed for stronger control.

\paragraph{From modular tuning to constrained learning.}
The current results also motivate a deeper paper-level pivot.
They do not show that modular design is misguided:
each module adds useful information.
What they do show is that separately trained and separately tuned modules do
not automatically solve the end-to-end answer-support-abstain problem.
The remaining gap is therefore best interpreted as an objective mismatch.
The system still learns mainly for answer extraction and only later receives
support and abstention corrections.
That observation motivates a revised next-stage comparison in which the current
modular pipeline is treated as the baseline and a support-constrained learning
objective is treated as the new method.
Under that framing, the earlier experiments become evidence for the limitation
of post-hoc modular control in this setting rather than failed attempts that
should be discarded.
The key methodological point is that this Stage 5 learner does not need to be
larger or more exotic than the modular baseline.
It can use the same encoder family and the same data while changing only the
training target:
answer extraction, support, and abstention are optimized jointly by a
shared-encoder multi-head model instead of in three separate stages.
That makes the next comparison much cleaner.
If the new learner wins, the gain is much easier to attribute to objective
design rather than to extra model capacity or a looser evaluation setup.

\paragraph{Next paper-facing extensions.}
With Stage 1, Stage 2, and Stage 3 now in place, the most natural paper
extensions are:
\begin{itemize}[leftmargin=1.25em]
  \item using the calibrated QA and support signals in a tighter fixed
  controller as the strongest modular baseline,
  \item comparing that modular baseline directly against a
  support-constrained learning objective trained under matched conditions,
  \item making the verifier less permissive on unsupported answers even after
  calibration if the modular baseline is still too weak,
  \item testing retrieval noise, and
  \item comparing adaptive balancing against simpler fixed baselines.
\end{itemize}
