Large language models can produce answers that are fluent and plausible while
still being unsupported by the available evidence.
For this project, the main concern is not hallucination as a vague catch-all
label, but unsupported answering delivered as if it were reliable.
That framing matters because improving average accuracy does not automatically
reduce overconfident unsupported outputs.

This paper takes a deliberately narrow starting point.
Instead of attempting to solve hallucination in general, we study a controlled
evidence-grounded question answering task in which a system must either answer
or abstain.
The current implementation uses SQuAD~2.0 \citep{rajpurkar2018know}
with one question and one evidence passage as input.
Stage 1 compares two training regimes:
\begin{enumerate}[leftmargin=1.25em]
  \item a baseline trained only on answerable examples, and
  \item an abstain-aware variant trained to treat unsupported cases as a
  first-class target behavior.
\end{enumerate}
Stage 2 then adds a verifier over question, passage, and predicted answer in
order to estimate whether an answer is actually supported before it is shown to
the user.
Stage 3 keeps the Stage 1 and Stage 2 models fixed and calibrates their score
outputs so that later control stages can act on better-behaved confidence
signals rather than raw overconfident values.
Taken together, these stages are no longer just incremental modules;
they are also diagnostic experiments.
Stage 1 shows that abstention supervision can change behavior but does not
fully preserve answer quality, Stage 2 shows that support is learnable but only
weakly changes end-to-end behavior when used as a permissive post-hoc gate,
and Stage 3 shows that both score streams are overconfident before
calibration.
This sequence suggests that the remaining bottleneck is not only model quality,
but also objective mismatch:
the system is still trained and selected mostly around answer correctness, then
patched afterward with separate support and abstention controls.

The central hypothesis is simple:
explicit abstention supervision should reduce unsupported answers without
causing an unacceptable collapse in answer quality on answerable questions.
That hypothesis is narrow enough to evaluate cleanly, but still meaningful as a
first step toward a broader program on grounded answering, support verification,
and calibrated uncertainty.
The next immediate question is whether a learned support signal can recover some
of the remaining quality-versus-grounding gap without undoing the abstention
gain from Stage 1.
The current paper now motivates a second hypothesis for the next phase:
post-hoc modular tuning has limited headroom in this setting, so a better
formulation should treat evidence-grounded QA as a support-constrained learning
problem rather than as answer prediction followed by separate filters.
Concretely, the next comparison keeps the same backbone and supervision source
but replaces the staged pipeline with a jointly trained answer-support-abstain
objective, so the paper can test whether the gain comes from changing the
learning target rather than from adding yet another module.

The intended application setting is also narrow.
\system\ is most naturally suited to bounded, evidence-grounded QA scenarios in
which answers should be tied to supplied documents and abstention is preferable
to unsupported guessing, such as internal knowledge assistants, support systems
over approved documentation, and policy or compliance QA.
It is not intended as a general solution for open-domain truthfulness, balanced
opinion generation, or fairness beyond the scope of the provided evidence.

This framing is also consistent with recent work arguing that hallucination
persists partly because modern evaluation pipelines reward guessing more than
appropriate uncertainty \citep{kalai2025why}.
Our contribution is narrower and more operational:
rather than explaining hallucination in general, Stage 1 tests whether an
evidence-grounded QA system can be trained and evaluated so that abstention is
treated as a useful target behavior instead of a failure by default, and
Stage 2 tests whether a learned support verifier can improve downstream
answer-versus-abstain decisions on top of that base system.
Stage 3 then asks whether those QA and support scores can be calibrated enough
to serve as credible control variables in the next stage.
If those stages succeed only partially, that is still informative:
it supports a reformulation in which future stages compare the current modular
pipeline against a direct support-constrained multi-head learner under matched
data, backbone, and evaluation conditions.

This LaTeX scaffold is intentionally written as a living project paper.
Some sections already reflect stable design decisions, while others are
structured to absorb final metrics, figures, and ablations once the
experimental results are locked in.
