Stage 1 uses a standard extractive question answering architecture with a
binary external decision:
the system must either return an answer span from the passage or output
\abstain.
The mechanism is intentionally simple.
Rather than introducing a separate verifier or controller at this stage, the
project asks whether abstention can already emerge from standard extractive QA
supervision when unanswerable cases are represented correctly during training.

\subsection{Encoder and QA heads}

Let the input be a question $q$ and a passage $p$.
After tokenization, the model receives a single packed sequence of the form
\begin{equation}
  x = [\mathrm{CLS}], q, [\mathrm{SEP}], p, [\mathrm{SEP}].
\end{equation}

The encoder produces contextual token representations
\begin{equation}
  h_1, h_2, \dots, h_T = \mathrm{Enc}_{\theta}(x),
\end{equation}
where $T$ is the tokenized sequence length.
In the default setup, the encoder is initialized from the DistilBERT checkpoint
\modelname.
The implementation goes through the Hugging Face \autoqaclass{} interface so
the backbone can be swapped later.

The question answering head is the standard two-logit extractive head.
For each token representation $h_t$, the model predicts:
\begin{align}
  s_t &= w_s^\top h_t + b_s, \\
  e_t &= w_e^\top h_t + b_e,
\end{align}
where $s_t$ is the start logit and $e_t$ is the end logit for token $t$.

For a candidate span $(i, j)$ with $i \leq j$, the raw span score is
\begin{equation}
  S(i, j) = s_i + e_j.
\end{equation}

This is the core deep learning mechanism of Stage 1:
the encoder learns a contextual representation of the question-passage pair,
and the two linear heads learn where the answer should begin and end.

\subsection{Training targets}

The project compares two training modes under the same architecture,
preprocessing, and optimization settings.

\paragraph{Baseline mode.}
The baseline is trained only on answerable examples.
This makes it a standard extractive QA model:
every training case has a supervised span target inside the passage.

\paragraph{Abstain-aware mode.}
The abstain-aware model is trained on both answerable and unanswerable
examples.
For answerable examples, the targets are the gold start and end token
positions, just as in the baseline.
For unanswerable examples, both the start and end targets are mapped to the
\(\mathrm{CLS}\) position.

This matters because it turns abstention into a first-class supervised outcome
without changing the external architecture.
The model is not given a separate abstention head.
Instead, it learns to use the score of the \(\mathrm{CLS}\) token as a null or
no-answer signal.

The optimization objective is the standard averaged cross-entropy loss over
start and end targets:
\begin{equation}
  \mathcal{L}
  =
  \frac{1}{2}
  \Big(
    \mathrm{CE}(s, y^{\mathrm{start}})
    +
    \mathrm{CE}(e, y^{\mathrm{end}})
  \Big),
\end{equation}
where \(y^{\mathrm{start}}\) and \(y^{\mathrm{end}}\) are the gold token
indices.
For unanswerable examples in the abstain-aware setting,
\begin{equation}
  y^{\mathrm{start}} = y^{\mathrm{end}} = c,
\end{equation}
with $c$ denoting the \(\mathrm{CLS}\) index.

\subsection{Long-context handling}

Because passages can exceed the maximum encoder length, Stage 1 uses the
standard sliding-window extractive QA preprocessing strategy.
Each example may be expanded into multiple overlapping features with a fixed
maximum sequence length and document stride.
During training, each feature receives span supervision relative to its own
window.
During evaluation, the pipeline keeps the metadata needed to map feature-level
predictions back to the original example.

This implementation detail is important in deep learning terms because the
model is trained over token windows rather than over raw passages directly.
The final example-level prediction is therefore an aggregation over
feature-level start and end logits.

\subsection{Inference mechanism}

At inference time, the model produces start and end logits for every feature of
an example.
Stage 1 then computes two competing quantities:
\begin{enumerate}[leftmargin=1.25em]
  \item the best valid answer span score across all candidate spans and all
  windows, and
  \item the best null score, defined as the sum of the start and end logits at
  the \(\mathrm{CLS}\) position.
\end{enumerate}

If the best span is \((i^\star, j^\star)\), then the best answer score is
\begin{equation}
  S_{\mathrm{span}}^\star
  =
  \max_{i, j \in \mathcal{V}} \; S(i, j),
\end{equation}
where \(\mathcal{V}\) contains only valid spans that satisfy the ordering and
maximum-answer-length constraints.

The null score is
\begin{equation}
  S_{\varnothing}
  =
  s_c + e_c,
\end{equation}
again using the \(\mathrm{CLS}\) position \(c\).

The baseline model always outputs the highest-scoring span.
The abstain-aware model compares the null score against the best span through a
margin:
\begin{equation}
  \Delta = S_{\varnothing} - S_{\mathrm{span}}^\star.
\end{equation}

The final decision rule is:
\begin{equation}
  \hat{y}
  =
  \begin{cases}
    \abstain, & \text{if } \Delta > \tau, \\
    \text{best span}, & \text{otherwise,}
  \end{cases}
\end{equation}
where \(\tau\) is a validation-selected threshold.

Conceptually, this means Stage 1 does not learn a symbolic rule for when to
refuse.
It learns a representation in which the no-answer state competes directly
against answer spans in the same scoring space.

\subsection{Threshold selection}

The abstention threshold is not fixed by hand in the default evaluation path.
Instead, the code performs a simple search over a bounded range on the
validation split.
For each candidate threshold, the system post-processes predictions and
measures Stage 1 metrics.
The current implementation selects the threshold by prioritizing abstain
\(F_1\), with answerable \(F_1\) used as a secondary tie-breaker.

This choice reflects the Stage 1 research question.
The project is not only asking whether the model can refuse more often; it is
asking whether it can reduce unsupported answers while preserving usefulness on
answerable examples.

\subsection{What Stage 1 proves}

From a deep learning perspective, Stage 1 is a representation and decision
control test.
It asks whether a standard extractive QA network, with no additional verifier
and no retrieval pipeline, can be pushed toward better answer-versus-abstain
behavior purely through:
\begin{itemize}[leftmargin=1.25em]
  \item different supervision over answerable versus unanswerable cases,
  \item a shared encoder and span head,
  \item the \(\mathrm{CLS}\)-based null mechanism, and
  \item thresholded post-processing over the learned logit space.
\end{itemize}

If this works, then later stages can build on a meaningful base mechanism
instead of introducing extra components before the basic abstention behavior is
understood.

\subsection{Stage 2 verifier and support gate}

Stage 2 keeps the Stage 1 abstain-aware QA model fixed and adds a second model
that judges whether a proposed answer is actually supported by the passage.
This makes the overall system modular rather than jointly end-to-end:
Stage 1 proposes an answer-or-\abstain\ decision, and Stage 2 decides whether a
predicted answer should be trusted.

\paragraph{Verification examples.}
The verifier is trained as a binary classification problem built from the same
underlying SQuAD~2.0 splits.
For each answerable training example, the pipeline creates one positive
verification record using a gold answer and one or more negative records using
sampled spans from the same passage that do not overlap strongly with the gold
answer.
For each unanswerable example, the pipeline creates only negative records by
sampling unsupported spans from the passage.
The result is a support-label dataset over tuples
\((q, p, a)\), where \(q\) is the question, \(p\) is the passage, and \(a\) is
a candidate answer.

\paragraph{Verifier input and model.}
The verifier formats the first sequence as a short hypothesis-like query,
\begin{equation}
  x^{(1)} = \texttt{Question: } q \; \texttt{\textbackslash n Answer: } a,
\end{equation}
and uses the passage as the second sequence,
\begin{equation}
  x^{(2)} = p.
\end{equation}
These paired sequences are fed into a standard transformer sequence classifier
initialized from the same \modelname\ checkpoint family as Stage 1, but with a
two-class support head.
If \(h_{\mathrm{CLS}}\) is the final pooled representation, the verifier
computes support logits
\begin{equation}
  z = W h_{\mathrm{CLS}} + b,
\end{equation}
followed by a softmax that estimates
\begin{equation}
  P_{\phi}(\text{supported} \mid q, p, a).
\end{equation}
Training uses standard cross-entropy over the binary support labels.

\paragraph{Inference and gating.}
At evaluation time, only examples that Stage 1 already labeled as
\texttt{answer} are sent to the verifier.
If the Stage 1 system abstains, the final decision remains \abstain.
If Stage 1 answers, Stage 2 attaches a support score
\(s_{\mathrm{sup}} \in [0, 1]\) and applies a second threshold:
\begin{equation}
  \hat{y}_{\mathrm{final}}
  =
  \begin{cases}
    \text{answer}, & \text{if } \hat{y}_{\mathrm{qa}}=\text{answer}
    \text{ and } s_{\mathrm{sup}} \ge \eta, \\
    \abstain, & \text{otherwise.}
  \end{cases}
\end{equation}
Here \(\eta\) is a validation-selected support threshold.
This means the Stage 2 system is a cascaded decision process:
the QA model first decides whether answering is plausible at all, and the
verifier then decides whether that proposed answer appears passage-supported.

\paragraph{Threshold search and Stage 2 objective.}
The support threshold is selected on the validation split by prioritizing
support \(F_1\), then gated overall \(F_1\), and then lower unsupported-answer
rate.
This reflects the Stage 2 question directly:
not only whether support is learnable, but whether a learned support signal can
improve the final answer-versus-\abstain\ decision.

\subsection{Revised problem formulation: support-constrained selective prediction}

The first three stages motivate a reformulation of the project objective.
Up to Stage 3, the system is intentionally modular:
the QA model is trained first, the verifier is trained second, and calibration
is fit after both models are frozen.
That modular structure is useful for diagnosis, but it also leaves the final
decision objective implicit.
The system is still learned mainly for answer prediction and then adjusted
afterward with verifier thresholds and calibrated score transformations.

The empirical pattern from Stages 1--3 suggests that this post-hoc modular path
has limited headroom in the current setting.
Stage 1 improves abstention behavior but sacrifices answerable-question
quality, Stage 2 learns support but changes end-to-end QA only marginally, and
Stage 3 improves score reliability without changing the underlying decision
objective.
This motivates a cleaner formulation in which evidence-grounded QA is treated
as a constrained selective prediction problem rather than as answer extraction
followed by separate filters.

Let the final system choose either an answer \(a\) or abstention
\(\varnothing\) for each question-passage pair \((q, p)\).
Define a task utility
\(U(\hat{y}, y, p)\)
that rewards a correct supported answer, penalizes an unsupported answer more
strongly, and penalizes unnecessary abstention more mildly.
The next-stage learning problem can then be written abstractly as
\begin{align}
  \max_{\theta} \quad & \mathbb{E}\left[ U(\hat{y}_{\theta}, y, p) \right] \\
  \text{s.t.} \quad &
  R_{\mathrm{unsup}}(\theta) \le \tau_{\mathrm{u}}, \\
  &
  R_{\mathrm{overabs}}(\theta) \le \tau_{\mathrm{a}},
\end{align}
where \(R_{\mathrm{unsup}}\) is unsupported-answer risk and
\(R_{\mathrm{overabs}}\) is unnecessary-abstention risk.
In words:
the system should maximize answer utility while staying within an explicit
groundedness budget.

A practical surrogate for this idea is a Lagrangian objective
\begin{equation}
  \mathcal{L}_{\mathrm{total}}
  =
  \mathcal{L}_{\mathrm{ans}}
  +
  \lambda_{\mathrm{u}} \mathcal{L}_{\mathrm{unsup}}
  +
  \lambda_{\mathrm{a}} \mathcal{L}_{\mathrm{overabs}},
\end{equation}
where \(\mathcal{L}_{\mathrm{ans}}\) is the standard answer loss,
\(\mathcal{L}_{\mathrm{unsup}}\) penalizes unsupported answering, and
\(\mathcal{L}_{\mathrm{overabs}}\) penalizes abstaining when a supported
correct answer was available.
This does not require a claim about a new universal learning law.
Its contribution is narrower and more defensible:
it makes the project's true target explicit instead of treating support and
abstention only as post-hoc corrections to answer accuracy.

\paragraph{A feasible deep-learning instantiation for Stage 5.}
The Stage 5 comparison does not require a new backbone or a new dataset.
It can be realized with the same encoder family already used in earlier
stages, but with a different training objective and a tighter coupling between
answer extraction, support, and abstention.
Concretely, let
\(x = [\mathrm{CLS}], q, [\mathrm{SEP}], p, [\mathrm{SEP}]\)
and let a shared encoder produce contextual states
\(h_1, \dots, h_T = \mathrm{Enc}_{\theta}(x)\).
Stage 5 then augments the standard extractive QA heads with two additional
binary decision heads so that answer extraction, support, and abstention are
learned jointly from the same representation.
The model exposes:
\begin{itemize}[leftmargin=1.25em]
  \item span logits
  \(z^{\mathrm{s}}_t = {w^{\mathrm{s}}}^\top h_t + b^{\mathrm{s}}\)
  and
  \(z^{\mathrm{e}}_t = {w^{\mathrm{e}}}^\top h_t + b^{\mathrm{e}}\)
  for answer start and end positions,
  \item an abstention logit
  \(z^{\varnothing} = {w^{\varnothing}}^\top h_{\mathrm{CLS}} + b^{\varnothing}\),
  and
  \item a support logit
  \(z^{\mathrm{sup}} = {w^{\mathrm{sup}}}^\top h_{\mathrm{CLS}} + b^{\mathrm{sup}}\).
\end{itemize}
Define
\(p_{\mathrm{abs}} = \sigma(z^{\varnothing})\)
and
\(p_{\mathrm{sup}} = \sigma(z^{\mathrm{sup}})\).
Let
\(y_{\mathrm{sup}} \in \{0,1\}\)
denote whether the gold answer is supported by the provided passage, and let
\(y_{\mathrm{keep}} \in \{0,1\}\)
denote whether the final system should emit an answer rather than
\abstain.
In the current single-passage setting, the two coincide:
\(y_{\mathrm{keep}} = y_{\mathrm{sup}} = 1\)
for answerable supported examples and
\(y_{\mathrm{keep}} = y_{\mathrm{sup}} = 0\)
for unanswerable or unsupported cases.
The model then defines a differentiable keep probability
\begin{equation}
  p_{\mathrm{keep}}
  =
  \left(1 - p_{\mathrm{abs}}\right) p_{\mathrm{sup}},
\end{equation}
which operationalizes the idea that an answer should survive only when the
model is simultaneously unwilling to abstain and confident that the answer is
supported.

A practical Stage 5 surrogate objective is then
\begin{align}
  \mathcal{L}_{\mathrm{stage5}}
  &=
  \mathcal{L}_{\mathrm{span}}
  +
  \lambda_{\mathrm{k}} \mathcal{L}_{\mathrm{keep}}
  +
  \lambda_{\mathrm{sup}} \mathcal{L}_{\mathrm{sup}}, \\
  \mathcal{L}_{\mathrm{span}}
  &=
  y_{\mathrm{keep}}
  \left[
    \mathrm{CE}(z^{\mathrm{s}}, i^{*})
    +
    \mathrm{CE}(z^{\mathrm{e}}, j^{*})
  \right], \\
  \mathcal{L}_{\mathrm{keep}}
  &=
  - \alpha y_{\mathrm{keep}} \log p_{\mathrm{keep}}
  - \beta \left(1-y_{\mathrm{keep}}\right)\log\!\left(1-p_{\mathrm{keep}}\right), \\
  \mathcal{L}_{\mathrm{sup}}
  &=
  \mathrm{BCE}\!\left(p_{\mathrm{sup}}, y_{\mathrm{sup}}\right),
\end{align}
with \(\beta > \alpha\) so that unsupported answering is penalized more
heavily than unnecessary abstention.
This is the deep-learning form of the constrained objective above:
\(\mathcal{L}_{\mathrm{span}}\) preserves standard extractive QA supervision,
\(\mathcal{L}_{\mathrm{keep}}\) learns the final answer-versus-\abstain\
decision under asymmetric cost, and
\(\mathcal{L}_{\mathrm{sup}}\) keeps the support head explicitly grounded.
The important difference from the modular pipeline is that these terms are
optimized jointly through the shared encoder instead of being trained in
separate stages and combined only at inference time.

This formulation is feasible in the current project for two reasons.
First, every term is differentiable with respect to the model outputs, so the
whole objective can be optimized end to end with standard gradient-based
training.
Second, the required supervision is already available or can be derived from
the current task setup:
gold spans provide \((i^{*}, j^{*})\), unanswerable examples supervise
abstention, and the support target can be constructed from the same
answerable-versus-unsupported labeling logic already used in Stages 2 and 3.
So Stage 5 is not a speculative architectural leap.
It is a controlled change in the learning objective that can be tested under
matched backbone, data, and evaluation conditions.

\subsection{Experimental design for the next comparison stage}

Under this revised formulation, the next comparison should be deliberately
simple and controlled.
The same backbone, dataset split, preprocessing, and core QA setup should be
kept fixed, while only the decision formulation changes.
The intended comparison is therefore:
\begin{enumerate}[leftmargin=1.25em]
  \item the current modular pipeline
  (Stage 1 abstention, Stage 2 support verification, and Stage 3 calibration),
  \item a fixed constrained controller over the calibrated signals, and
  \item a direct support-constrained learner with the same encoder family and
  preprocessing pipeline, trained under matched conditions.
\end{enumerate}
This comparison design is important because it isolates the real question:
whether the main remaining gain comes from further module tuning, or from
changing the learning objective itself.

\subsection{Stage 3 confidence calibration}

Stage 3 does not retrain the Stage 1 QA model or the Stage 2 verifier.
Instead, it treats both of them as fixed score generators and asks whether
their confidence values can be made more trustworthy through lightweight
post-hoc calibration.
The goal is not yet to change the final controller.
The goal is to produce calibrated decision signals that later stages can trust
when deciding whether to answer, abstain, or suppress a weakly supported
answer.

\paragraph{QA decision calibration.}
For the Stage 1 abstain-aware model, Stage 3 reuses the same
validation-selected abstention threshold \(\tau\) and constructs a scalar QA
decision score from the margin between the null score and the best answer-span
score.
If the QA system predicts \abstain, the score is defined as
\begin{equation}
  z_{\mathrm{qa}}
  =
  \Delta - \tau,
\end{equation}
and if it predicts an answer, the sign is flipped so that larger values always
mean \emph{more confidence that the final QA decision was correct}:
\begin{equation}
  z_{\mathrm{qa}}
  =
  \tau - \Delta.
\end{equation}
The binary QA calibration label is 1 if the final Stage 1 decision was correct
under the decision-aware evaluation rule and 0 otherwise.

\paragraph{Support-score calibration.}
For Stage 2, only QA predictions that remain \texttt{answer} are passed to the
support verifier, just as in the original Stage 2 pipeline.
The verifier produces a support probability
\(
  p_{\mathrm{sup}} = P_{\phi}(\text{supported} \mid q,p,a)
\).
Stage 3 first maps that probability back to a logit-like score through the
inverse sigmoid:
\begin{equation}
  z_{\mathrm{sup}}
  =
  \log\frac{p_{\mathrm{sup}}}{1 - p_{\mathrm{sup}}},
\end{equation}
and then calibrates this score against a binary target indicating whether the
predicted answer actually matches a gold answer strongly enough to count as
supported.

\paragraph{Temperature scaling.}
Both score streams are calibrated separately with scalar temperature scaling.
Given a raw score \(z\) and a learned temperature \(T > 0\), the calibrated
probability is
\begin{equation}
  \tilde{p}
  =
  \sigma\!\left(\frac{z}{T}\right),
\end{equation}
where \(\sigma(\cdot)\) is the sigmoid function.
The implementation searches over a bounded grid of temperatures on the
validation split and picks the value that minimizes the Brier score, with
expected calibration error (ECE) used as the tie-breaker.
This is a deliberately simple post-hoc mechanism:
the underlying QA and verifier parameters stay fixed, and only the sharpness of
their confidence mapping changes.

\paragraph{Stage 3 metrics and artifacts.}
Stage 3 reports raw-versus-calibrated metrics for both QA and support scores on
the held-out development split.
The main reported quantities are ECE, adaptive ECE, maximum calibration error
(MCE), Brier score, a threshold-gap measure, score-label correlation, and
reliability diagrams.
Because the same Stage 1 QA threshold and Stage 2 support threshold are kept
fixed, Stage 3 should be read as a calibration study rather than as a new
end-to-end controller.
Its success criterion is narrower:
reduce overconfidence and make the score distributions more faithful to
empirical correctness so that later control stages have a better starting
signal.
In deep learning terms, Stage 2 tests whether a second learned representation
space over \((q, p, a)\) can act as a useful post-hoc controller on top of the
Stage 1 answer space.
